Executive Constraints and Civil Conflict Onset

### 2020-05-26 paper.Rmd
Add text references to manuscript

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: "`r format(Sys.time(), '%Y-%m-%d')`"
suppress-bibliography: false
bibliography: library.bib
csl: assets/american-political-science-association.csl
link-citations: true
documentclass: scrreprt
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian structural equation model with a measurement portion estimating
executive constraints as a latent construct and a multilevel logistic
regression relating the latent construct to the observed outcome,
civil conflict.

@@

[^model]: The model code is written in Stan [@Carpenter2017] and was
    run using the `CmdStan` interface version 2.22.1 [@Stan2020] with
    four parallel chains at 2000 iterations each with a warmup
    of 1000. Convergence was assessed according to the potential scale
    reduction factor $\widehat{R}$ and effective sample size (ESS)
    [@Vehtari2019] in addition to standard Stan diagnostics. All
    pre/post-processing was done using R 3.6.2 [@Core2019]. The Stan
    model code is available in [Appendix A](#appendix-a) and the full
    project in the form of a reproducible pipeline, including the
    Rmarkdown source for this manuscript, is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis).
    Finally, a simulation analysis assessing the model and it's
    ability to recover simulated true values is available in [Appendix
    B](#appendix-b).

@@

```{r fig.cap = "\\label{fig:onsets} Proportion of countries experiencing ongoing civil conflicts and new onsets"}

@@

For the purposes of this paper I estimate an index measuring
executive constraints from a factor analysis of multiple manifest
variables drawn from the V-Dem dataset. The aim is to capture the
extent to which the executive is *de facto* constrained by other
government institutions --- more specifically, the legislature,
judiciary, and independent investigative bodies --- rather than the
*de jure* powers of any one institution *vis-à-vis* the executive, and
to do so using a latent variable aggregation that avoids the arbitrary
weighting of linear additive indices.[^additive] The aggregation
method is largely similar to the accountability index created by
@Luhrmann2017 and the presidentialism index created by @Sigman2018. By
creating a separate index, however, I am able to adjust the list of
included manifest variables as well as embed the factor analysis
within the larger regression model such that the uncertainty from the
latent factor propagates as a result of the estimation process.

@@

@@

kable(q, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "Descriptive statistics for manifest variables") %>%
    kable_styling(latex_options = "striped")

@@

```{r fig.height = 5, fig.width = 7, fig.cap = "\\label{fig:cor} Correlation Heatmap"}
m <- cor(manifest, use = "pairwise.complete.obs") %>%
    signif(3)

@@

@@

kable(q, "latex", align = c("l", "r", "r", "r", "r", "r"),
      booktabs = T, linesep = "", escape = T,
      caption = "Descriptive statistics for regression covariates") %>%
    kable_styling(latex_options = "striped")

@@

```{r fig.height = 4, fig.cap = "\\label{fig:fpars} Measurement Model Parameter Estimates"}
load("./posteriors/full_model/input.RData")
fa <- read_post("./posteriors/full_model/fa_posteriors.csv")

@@

@@

kable(fa.mm, "latex", align = "c", booktabs = T, linesep = "", escape = F,
    kable_styling(latex_options = c("striped", "scale_down"), full_width = F,
                  protect_latex = F)
```

@@

```{r fig.width = 6, fig.height = 2, fig.cap = "\\label{fig:bicam} Posterior density plots for executive constraints"}
theta <- extract(fa, "theta")

@@

```{r fig.cap = "\\label{fig:fa-line} Posterior shrinkage for \\textit{v2juncind}"}
err <- read_post("posteriors/full_model/err_posteriors.csv")

@@

```{r fig.cap = "\\label{fig:err} Posterior CI and point estimates vs. observed error"}
p1 <- ggplot(nonlg_est, aes(se, interval)) +
    geom_point(col = "#56B4E9", alpha = 0.2, shape = 1) +
    theme_tufte() +
    ylab("Posterior CI Range") +
    xlab("Observed Error")

@@

```{r fig.width = 7, fig.height = 5, fig.cap = "\\label{fig:theta} Executive Constraints Over Time"}
theta_summarised <- post_summarise(theta, probs = c(0.10, 0.5, 0.90)) %>%
    rename(codelow = `10%`, median = `50%`, codehigh = `90%`)

@@

```{r results = "hide", fig.width = 8, fig.cap = "\\label{fig:map} Executive Constraints - 2012. Grey indicates countries missing from the dataset."}
map_data <- st_read("data/raw/cshapes_0.6/cshapes.shp") %>%
    filter(GWEYEAR == max(GWEYEAR)) %>%
    select(country_name = CNTRY_NAME, gwid = GWCODE)

@@

```{r fig.cap = "\\label{fig:tie} Executive Constraints 2000 vs 2010"}
targets <- c("Algeria", "Argentina", "Armenia", "Chile", "Colombia",
             "Egypt", "Georgia", "Ghana", "Hungary", "Indonesia",
             "Iran", "Israel", "Japan", "Kenya", "Mexico", "Poland",
             "Portugal", "Romania", "Rwanda", "Turkey", "Tunisia",
             "United States of America", "Zimbabwe")

@@

@@

kable(beta_s, "latex", align = c("c", "r", "r"), booktabs = T,
      linesep = "", escape = F,
      caption = "Posterior credible intervals for regression slopes") %>%
    kable_styling(latex_options = "striped", "scale_down")
```

```{r fig.cap = "\\label{fig:reg} Regression parameters"}
plot_pars(beta, hline = T)
```

@@

```{r fig.cap = "\\label{fig:year} Log-odds year intercepts - $80\\%$ CIs"}
Z_year <- extract(reg, "Z_year")
colnames(Z_year) <- { unique(final.df$year) + 1 } %>% sort

@@

```{r, fig.cap = "\\label{fig:country} Log-odds country intercepts - $80\\%$ CIs"}

@@

```{r, fig.cap = "\\label{fig:peace} Log-odds peace year intercepts - $80\\%$ CIs"}
f <- extract(reg, "f")
colnames(f) <- unique(final.df$lpeace_yrs) %>% sort

@@

```{r fig.cap = "\\label{fig:int} Interaction between executive constraints and GDP per capita with 80 percent CIs"}
# Interaction plot
medians <- apply(theta, 2, quantile, probs = 0.5)
lower_bound <- min(medians) %>% round(digits = 1)
upper_bound <- max(medians) %>% round(digits = 1)

@@

```{r fig.cap="\\label{fig:gdp} Difference in predicted probabilities between the lowest and highest level of executive constraints for different values of GDP per capita"}
rgdpepc <- seq(min(X[, "rgdpepc"]), state[1], 0.01)
df <- mclapply(rgdpepc, function(y) {
    low <- boot::inv.logit(fn(lower_bound, y)) * 100
    high <- boot::inv.logit(fn(upper_bound, y)) * 100

@@

```{r fig.height = 2, fig.cap = "\\label{fig:phat} Density plot for predicted probabilities"}
extra <- read_post("posteriors/full_model/extra_posteriors.csv")
p_hat <- extract(extra, "p_hat")

@@

@@

kable(df, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "Summarized posterior predicted probabilities")
```

@@

kable(m, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "Confusion matrix with a 0.2 cutoff")
```

@@

```{r fig.width = 8, fig.cap = "\\label{fig:roc} ROC & Precision-Recall Plots"}

autoplot(x, curvetype = c("ROC", "PRC"))
```

>>>>

---
title: Executive Constraints and Civil Conflict Onset
author: J.L.A. Krusell
date: "`r format(Sys.time(), '%Y-%m-%d')`"
suppress-bibliography: false
bibliography: library.bib
csl: assets/american-political-science-association.csl
link-citations: true
documentclass: scrreprt
fontsize: 11pt
linestretch: 1.5
header-includes:
    - \usepackage{amsmath}
    - \usepackage{etoolbox}
    - \usepackage{lualatex-math}
    - \usepackage{unicode-math}
    - \appto\Shaded{\singlespacing}
    - \BeforeBeginEnvironment{align*}{\begin{singlespace}\vspace*{-\baselineskip}}
    - \AfterEndEnvironment{align*}{\end{singlespace}\noindent\ignorespaces}
    - \counterwithout{table}{chapter}
    - \counterwithout{figure}{chapter}
output:
    pdf_document:
        fig_caption: true
        keep_tex: true
        latex_engine: lualatex
        toc: true
        toc_depth: 2
        pandoc_args: ["--syntax-definition", "assets/stan.xml",
                      "--highlight-style", "kate",
                      "--lua-filter", "utils/text_refs/text_refs.lua"]
    html_document:
        smart: true
        toc: true
abstract: Lorem ipsum
---

@@

To examine the relationship between civil conflict onset and executive
constraints, I construct a dataset of relevant variables and run a
bayesian latent variable model with a measurement portion
estimating executive constraints as a latent construct and a
multilevel logistic regression relating the latent construct to the
observed outcome, civil conflict.

@@

[^model]: The model code is written in Stan [@Carpenter2017] and was
    run using the `CmdStan` interface version 2.23.0 [@Stan2020] with
    four parallel chains at 2000 iterations each with a warmup
    of 1000. Convergence was assessed according to the potential scale
    reduction factor $\widehat{R}$ and effective sample size (ESS)
    [@Vehtari2019] in addition to standard Stan diagnostics. All
    pre/post-processing was done using R 3.6.2 [@Core2019]. The Stan
    model code is available in [Appendix A](#appendix-a) and the full
    project in the form of a reproducible pipeline, including the
    Rmarkdown source for this manuscript, is available at
    [https://github.com/jsks/ma-thesis](https://github.com/jsks/ma-thesis).
    Finally, a simulation analysis assessing the model and it's
    ability to recover simulated true values is available in [Appendix
    B](#appendix-b).

@@

(ref:onsets) Proportions of countries per year experiencing at least
    one ongoing civil conflict and at least one new onset.

```{r fig.cap = "\\label{fig:onsets} (ref:onsets)"}

@@

For the purposes of this paper I estimate an index measuring executive
constraints from a factor analysis of multiple manifest variables
drawn from the V-Dem dataset. The aim is to capture the extent to
which the executive is *de facto* constrained by other government
institutions --- more specifically, the legislature, judiciary, and
independent investigative bodies --- rather than the *de jure* powers
of any one institution *vis-à-vis* the executive, and to do so using a
latent variable aggregation that avoids the arbitrary weighting of
linear additive indices.[^additive] The aggregation method is largely
similar to the accountability index created by @Luhrmann2017 and the
presidentialism index created by @Sigman2018 already available in the
V-Dem dataset. By creating a separate index, however, I am able to
adjust the list of included manifest variables as well as embed the
factor analysis within the larger regression model such that the
uncertainty from the latent factor propagates as a result of the
estimation process.

@@

(ref:fa-sum) Descriptive statistics for manifest variables

@@

kable(q, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "(ref:fa-sum)") %>%
    kable_styling(latex_options = "striped")

@@

(ref:cor) Correlation heatmap of manifest variables

```{r fig.height = 5, fig.width = 7, fig.cap = "\\label{fig:cor} (ref:cor)"}
m <- cor(manifest, use = "pairwise.complete.obs") %>%
    signif(3)

@@

(ref:reg-stats) Descriptive statistics for regression covariates

@@

kable(q, "latex", align = c("l", "r", "r", "r", "r", "r"),
      booktabs = T, linesep = "", escape = T,
      caption = "(ref:reg-stats)") %>%
    kable_styling(latex_options = "striped")

@@

(ref:fpars) Executive constraints measurement model parameter
    estimates with 80 percent posterior credible intervals.

```{r fig.height = 4, fig.cap = "\\label{fig:fpars} (ref:fpars)"}
load("./posteriors/full_model/input.RData")
fa <- read_post("./posteriors/full_model/fa_posteriors.csv")

@@

(ref:fa)

Executive constraints measurement model parameter
    estimates with 80 percent posterior credible intervals

@@

kable(fa.mm, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "(ref:fa)") %>%
    kable_styling(latex_options = c("striped", "scale_down"), full_width = F,
                  protect_latex = F)
```

@@

(ref:bicam) Posterior density plots for the executive constraints
    measure, $\theta$.

```{r fig.width = 6, fig.height = 2, fig.cap = "\\label{fig:bicam} (ref:bicam)"}
theta <- extract(fa, "theta")

@@

(ref:fa-line) Posterior shrinkage for *v2juncind*. The observed point
    estimates are the data points reported by V-Dem.

```{r fig.cap = "\\label{fig:fa-line} (ref:fa-line)"}
err <- read_post("posteriors/full_model/err_posteriors.csv")

@@

(ref:err) On the left, the posterior uncertainty for *v2juncind* as
    estimated by the model plotted against the observed error reported
    by V-Dem as *v2juncind_sd*. On the right, the difference between
    the point estimates estimated by the model and the point estimates
    reported by V-Dem versus the observed error for *v2juncind*.

```{r fig.cap = "\\label{fig:err} (ref:err)"}
p1 <- ggplot(nonlg_est, aes(se, interval)) +
    geom_point(col = "#56B4E9", alpha = 0.2, shape = 1) +
    theme_tufte() +
    ylab("Posterior CI Range") +
    xlab("Observed Error")

@@

(ref:theta) Executive constraints over time for several selected
    countries.

```{r fig.width = 7, fig.height = 5, fig.cap = "\\label{fig:theta} (ref:theta)"}
theta_summarised <- post_summarise(theta, probs = c(0.10, 0.5, 0.90)) %>%
    rename(codelow = `10%`, median = `50%`, codehigh = `90%`)

@@

(ref:map) Executive constraints in 2012 for all countries present in
    the dataset. Missingness indicated by the color grey.

```{r results = "hide", fig.width = 8, fig.cap = "\\label{fig:map} (ref:map)"}
map_data <- st_read("data/raw/cshapes_0.6/cshapes.shp") %>%
    filter(GWEYEAR == max(GWEYEAR)) %>%
    select(country_name = CNTRY_NAME, gwid = GWCODE)

@@

(ref:cap) Executive constraints in 2000 versus 2010 for several
    selected countries.

```{r fig.cap = "\\label{fig:tie} (ref:cap)"}
targets <- c("Algeria", "Argentina", "Armenia", "Chile", "Colombia",
             "Egypt", "Georgia", "Ghana", "Hungary", "Indonesia",
             "Iran", "Israel", "Japan", "Kenya", "Mexico", "Poland",
             "Portugal", "Romania", "Rwanda", "Turkey", "Tunisia",
             "United States of America", "Zimbabwe")

@@

(ref:reg) 80 percent posterior credible intervals for regression
    slopes

@@

kable(beta_s, "latex", align = c("c", "r", "r"), booktabs = T,
      linesep = "", escape = F,
      caption = "(ref:reg)") %>%
    kable_styling(latex_options = "striped", "scale_down")
```

(ref:fig-reg) Regression slopes with 80 percent posterior credible
    intervals.

```{r fig.cap = "\\label{fig:reg} (ref:fig-reg)"}
plot_pars(beta, hline = T)
```

@@

(ref:year) Log-odds year intercepts with 80 percent credible intervals

```{r fig.cap = "\\label{fig:year} (ref:year)"}
Z_year <- extract(reg, "Z_year")
colnames(Z_year) <- { unique(final.df$year) + 1 } %>% sort

@@

(ref:country) Log-odds country intercepts with 80 percent credible
    intervals

```{r, fig.cap = "\\label{fig:country} (ref:country)"}

@@

(ref:peace) Log-odds peace year intercepts with 80 percent credible
    intervals

```{r, fig.cap = "\\label{fig:peace} (ref:peace)"}
f <- extract(reg, "f")
colnames(f) <- unique(final.df$lpeace_yrs) %>% sort

@@

(ref:int) Interaction effect between executive constraints and GDP per
    capita with 80 percent credible intervals.

```{r fig.cap = "\\label{fig:int} (ref:int)"}
# Interaction plot
medians <- apply(theta, 2, quantile, probs = 0.5)
lower_bound <- min(medians) %>% round(digits = 1)
upper_bound <- max(medians) %>% round(digits = 1)

@@

(ref:gdp) Difference in predicted probabilities between the lowest and
    highest level of executive constraints for different values of GDP
    per capita.

```{r fig.cap="\\label{fig:gdp} (ref:gdp)"}
rgdpepc <- seq(min(X[, "rgdpepc"]), state[1], 0.01)
df <- mclapply(rgdpepc, function(y) {
    low <- boot::inv.logit(fn(lower_bound, y)) * 100
    high <- boot::inv.logit(fn(upper_bound, y)) * 100

@@

(ref:phat) Density plot of predicted probabilities for actual onset
    country-years and no onset country-years.

```{r fig.height = 2, fig.cap = "\\label{fig:phat} (ref:phat)"}
extra <- read_post("posteriors/full_model/extra_posteriors.csv")
p_hat <- extract(extra, "p_hat")

@@

(ref:psum) Summarized posterior predicted probabilities

@@

kable(df, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "(ref:psum)")
```

(ref:confusion) Confusion matrix with a 0.168 cutoff

@@

kable(m, "latex", align = "c", booktabs = T, linesep = "", escape = F,
      caption = "(ref:confusion)")
```

@@

(ref:roc) ROC & Precision-Recall plots

```{r fig.width = 8, fig.cap = "\\label{fig:roc} (ref:roc)"}
autoplot(x, curvetype = c("ROC", "PRC"))
```
